---
title: "Assignment 10: Data Scraping"
author: "Laura Piccirillo"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 loading packages and checking working directory
#install.packages("tidyverse")
#install.packages("rvest")
#install.packages("ggplot2")
#install.packages("here")
library(tidyverse)
library(rvest)
library(ggplot2)
library(here)

getwd()
here()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 reading in the scraping html
webpage <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 scraping website values
water_system_name <- webpage %>%
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>%
  html_text()
water_system_name

PWSID <- webpage %>%
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>%
  html_text()
PWSID
  
ownership <- webpage %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()
ownership

MGD <- webpage %>%
  html_nodes('th~ td+ td , th~ td+ td') %>%
  html_text()
MGD

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in date format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 creating data frame
year <- '2024'
water_data <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                         "Water system name" = water_system_name,
                         "PWSID" = PWSID,
                         "Ownership" = ownership,
                         "Maximum Day Use (MGD)" = as.numeric(MGD)) %>%
  mutate(Date = my(paste(Month, "-", year)))

#5 line plot
ggplot(water_data, aes(Date, Maximum.Day.Use..MGD.)) + geom_line() +
  ggtitle('Monthly maximum daily water use') + 
  labs(subtitle = "Durham Municipality, 2024", y = "Maximum Day Use (MGD)") + 
  theme_bw()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. creating function to scrape website
scrape_website <- function(PWSID, year){
  website <- read_html(paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
    PWSID, '&year=', year))
  
  water_system_name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  MGD_tag <- 'th~ td+ td , th~ td+ td'
  
  water_system_name <- website %>% html_nodes(water_system_name_tag) %>% 
    html_text()
  PWSID <- website %>% html_nodes(PWSID_tag) %>% html_text()
  ownership <- website %>% html_nodes(ownership_tag) %>% html_text()
  MGD <- website %>% html_nodes(MGD_tag) %>% html_text()
  
  function_data <- data.frame("Month" = c(1,5,9,2,6,10,3,7,11,4,8,12),
                              "Year" = rep(year,12),
                         "Water system name" = water_system_name,
                         "PWSID" = PWSID,
                         "Ownership" = ownership,
                         "Maximum Day Use (MGD)" = as.numeric(MGD)) %>%
  mutate(Date = my(paste(Month, "-", year)))
    
  
  return(function_data)
  
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 scraping data for 2020, plotting it
data_2020 <- scrape_website('03-32-010', 2020)

glimpse(data_2020)

ggplot(data_2020, aes(Date, Maximum.Day.Use..MGD.)) + geom_line() +
  ggtitle('Monthly maximum daily water use') + 
  labs(subtitle = "Durham Municipality, 2020", y = "Maximum Day Use (MGD)") + 
  theme_bw()


```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 scraping data from Cary for 2020 and combining with Durham 2020

cary_2020 <- scrape_website('03-92-020', 2020)
all_2020 <- rbind(data_2020, cary_2020)

ggplot(all_2020, aes(Date, Maximum.Day.Use..MGD., colour = Water.system.name)) +
  geom_line(linewidth = 1.5) +
  ggtitle('Monthly maximum daily water use') + 
  labs(subtitle = "Durham and Cary, 2020", y = "Maximum Day Use (MGD)", 
       color = "Municipality") + 
  theme_bw()

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 Cary data from 2018 to 2023
the_years <- c(2018:2023)
all_Cary <- map2('03-92-020', the_years, scrape_website) %>%
  bind_rows()

ggplot(all_Cary, aes(Date, Maximum.Day.Use..MGD.)) + geom_line() + 
  geom_smooth(method = "loess", se = F) +
  ggtitle('Monthly maximum daily water use') + 
  labs(subtitle = "Cary, 2018-2023", y = "Maximum Day Use (MGD)") + 
  theme_bw()

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?

> Answer: Cary is showing a small trend overtime of increasing water usage from 2018 to 2023.
