---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Laura Piccirillo"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2, warning=FALSE}
#1 setting up session
#install.packages("tidyverse")
#install.packages("agricolae")
#install.packages("here")
#install.packages("corrplot")
library(tidyverse)
library(agricolae)
library(here)
library(corrplot)

getwd()
here()

Lake <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)

Lake$sampledate <- mdy(Lake$sampledate)

#2 building plot theme
mytheme <- theme_bw() +
  theme(axis.text = element_text(color = "black"), legend = "top")

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: Yes, mean lake temperature in July changes with depth across all lakes.
H0: Mean lake temperature in July does not change with depth across all lakes.
Ha: Mean lake temperature in July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, warning=FALSE}
#4 wrangle dataset

Lake_wrangled <- Lake %>%
  mutate(month = month(sampledate)) %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()
  
#5 visualize depth vs lake temperature
ggplot(Lake_wrangled, aes(depth,temperature_C)) + geom_point() + 
  mytheme + labs(y = "Temperature (Celsius)", x = "Depth (m)") + 
  geom_smooth(method = lm) + ylim(0, 35)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest anything about the linearity of this trend?

> Answer: Depth and temperature have a negative correlation due to decreasing temperatures with increasing depth. The distribution of points suggest the trend may not best be described as linear, as the temperature values above 8-9m in depth remain steady. Also, there is a wide spread of temperatures for each depth, either representing the different lakes or different years included in this study. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7 linear regression of temperature by depth

Lake_lm <- lm(data = Lake_wrangled, temperature_C ~ depth)
summary(Lake_lm)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: There is a significant negative correlation between temperature and depth, and about 74% of the variability in temperature is explained by depth, based on the R-squared value of 0.7387. This is based on 9726 degrees of freedom. The significance of this correlation is confirmed by the p-value of < 2.2e-16. For every 1m change in depth, temperature is expected to change by 1.95 degrees C, based on the depth coefficient of the linear model. Since these variables are negatively correlated, a 1m increase in depth will lead to a temperature decrease of 1.95 degrees C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC
Lake_wrangled$year4 <- as.numeric(Lake_wrangled$year4)
Lake_wrangled$daynum <- as.numeric(Lake_wrangled$daynum)

Lake_a_wrangled <- Lake_wrangled %>%
  select(year4, daynum, depth, temperature_C)

Lake_wrangled_corr <- cor(Lake_a_wrangled)
corrplot(Lake_wrangled_corr, method = "ellipse")

Temp_AIC <- lm(data = Lake_a_wrangled, temperature_C ~ year4 + daynum + depth)
step(Temp_AIC)

#10 multiple regression
multi_Lake <- lm(data = Lake_a_wrangled, temperature_C ~ year4 + daynum + depth)
summary(multi_Lake)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: All explanatory variables from the subset of the Lake data (year, daynum, depth) are suggested to be used in our multiple regression based on the AIC method, as it showed higher AIC values if you subtract any of these variables from the model. This multiple regression model explains 74% of the observed variance in temperature. This is a slight improvement over the model with only depth as an explanatory variable (R-squared of 0.7412 vs 0.7387 - depth only)

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Lake_temp_sum <- Lake_wrangled %>%
  group_by(lakename) %>%
  summarise(temperature_C = mean(temperature_C))

Lake_anova <- aov(data = Lake_temp_sum, temperature_C ~ lakename)
summary(Lake_anova)

Lake_lm_anova <- lm(data = Lake_temp_sum, temperature_C ~ lakename)
summary(Lake_lm_anova)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
